---
output: html_document
---
[comment]: <> (Rossano Marchetti Ramos / 10/10/2016)

# Reproducible Research - Johns Hopkins / Coursera
# Peer-graded Assignment: Course Project 1

## 1. Code for reading in the dataset and/or processing the data

```{r inicial}
################################################################################################
#
#           INSTRUCTIONS
#
#-----------------------------------------------------------------------------------------------
#
#    Dataset: Activity monitoring data [52K]
# 
# The variables included in this dataset are:
#
#    steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
#    date: The date on which the measurement was taken in YYYY-MM-DD format
#    interval: Identifier for the 5-minute interval in which measurement was taken
#
################################################################################################

# globl options
knitr::opts_chunk$set(echo=TRUE)

# -----------------------------------------------------------------
# 1. Code for reading in the dataset and/or processing the data
# -----------------------------------------------------------------
files <- dir()

#download and open file
if (sum(grepl("activity.csv", files)) == 0) {
    # .csv file hadn't been downloaded before
    if (sum(grepl("activity.zip", files))  == 0) {
        # .zip hadn't been downloaded before
        download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
                      destfile = "activity.zip")
        unzip("activity.zip")
    }
}
rm(files)

activNAs <- read.csv("activity.csv")
# transform 'date' class
activNAs$date <- as.Date(activNAs$date,"%Y-%m-%d")

# remove NA's. There is NA's in the column 'steps'
activ <- activNAs[complete.cases(activNAs),]
```

## 2. Total number of steps taken each day

```{r Histogram, fig.height=3.5, fig.width=5.5, dpi = 300}
# -----------------------------------------------------------------
# 2. Histogram of the total number of steps taken each day
# -----------------------------------------------------------------
d01 <- data.frame(date  = unique(activ$date), 
                  steps =  NA)
d01$steps <- sapply(d01$date, function(i)
    sum(activ$steps[activ$date == i]))

hist(d01$steps, main = "Histogram of the total number of steps taken each day",
     xlab = "number of steps", cex.main = 1)
```

## 3. The Mean and median number of steps taken each day are, respectively, `r format(mean(d01$steps), digits = 1)` and `r median(d01$steps)`.

## 4. Time series plot of the average number of steps taken

```{r TimeSeries, fig.height=3.5, fig.width=5.5, dpi = 300}
# -----------------------------------------------------------------
# 4. 
# -----------------------------------------------------------------
d02 <- data.frame(interval = unique(activ$interval), 
                  av.steps = NA,
                  seq.interval = 0:287)
# calculate mean steps in each interval
d02$av.steps <- sapply(d02$interval, function(i)
    mean(activ$steps[activ$interval == i]))
# seq of 5 min intervals: from 0 to 287 == 0:00 to 23:55
d02$hour <- d02$seq.interval*5/60

plot(av.steps ~ hour, type = "l", data = d02, 
     main = "Average number of steps taken during the day\n(5 min interval)",
     xlab = "time of the day (in hours)", ylab = "steps (average number)",
     cex.main = 1)
```

```{r MaxSteps}
# -----------------------------------------------------------------
# 5. 
# -----------------------------------------------------------------
max.steps <- which(max(d02$av.steps) == d02$av.steps)

# function that transforms dec hour in %H:%M format
# http://stackoverflow.com/questions/14483629/how-convert-decimal-to-posix-time
to.time <- function (x) {
    paste(floor(x), 
          round((x-floor(x))*60), 
          sep=":")
}
```

## 5. The 5-minute interval that, on average, contains the maximum number of steps starts at  `r to.time(d02$hour[max.steps])`.

## 6. Code to describe and show a strategy for imputing missing data
```{r ImputingNAs}
# -----------------------------------------------------------------
# 6. 
# -----------------------------------------------------------------
# creat a new variable and imput # of steps using
# the average steps 
activNAs$steps.new <- activNAs$steps
NAs.founded <- which(is.na(activNAs$steps))
activNAs$steps.new[NAs.founded] <- 
    sapply(activNAs$interval[NAs.founded], function(interval.NA) {
        d02$av.steps[d02$interval == interval.NA]})
    
# 
d03 <- data.frame(date      = unique(activ$date), 
                  steps.new =  NA)
d03$steps.new <- sapply(d03$date, function(i)
    sum(activNAs$steps.new[activNAs$date == i]))
```

# 7. Histogram of the total number of steps taken each day after missing values are imputed.

```{r ImputingNAs_results, fig.height=3.5, fig.width=5.5, dpi = 300}
# -----------------------------------------------------------------
# 7. 
# -----------------------------------------------------------------
hist(d03$steps.new, main = "Histogram of the total number of steps taken each day\n (imputed NAs values)", xlab = "number of steps",cex.main = 1)
```

After imputing values for NAs, the mean and median number of steps taken each day are, respectively, `r format(mean(d03$steps.new), digits = 1)` and `r format(median(d03$steps.new), digits = 1)` (without NAs the values were `r format(mean(d01$steps), digits = 1)` and `r median(d01$steps)`).

# 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r PanelPlot, fig.height=8, fig.width=7.5, dpi = 300}
# -----------------------------------------------------------------
# 8. 
# -----------------------------------------------------------------
activ$type <- as.factor(
        ifelse(weekdays(activ$date) %in% c("domingo", "sábado"), 
           "weekend", "weekday"))

dweekday <- data.frame(interval = unique(activ$interval), 
                       av.steps = NA,
                       seq.interval = 0:287)
dweekend <- data.frame(interval = unique(activ$interval), 
                       av.steps = NA,
                       seq.interval = 0:287)

# calculate mean steps in each interval
dweekday$av.steps <- sapply(dweekday$interval, function(i)
    mean(activ$steps[activ$type == "weekday" & activ$interval == i]))
dweekend$av.steps <- sapply(dweekend$interval, function(i)
    mean(activ$steps[activ$type == "weekend" & activ$interval == i]))

# seq of 5 min intervals: from 0 to 287 == 0:00 to 23:55
dweekday$hour <- dweekday$hour <- dweekday$seq.interval*5/60
dweekend$hour <- dweekend$hour <- dweekend$seq.interval*5/60

par(mfrow = c(2, 1), mar = c(5.1, 4.1, 1, 2.1 ))
plot(av.steps ~ hour, type = "l", data = dweekday, 
     xlab = "time of the day (in hours)", ylab = "steps (average number)")
mtext("Weekday", side = 3, line = -1)
plot(av.steps ~ hour, type = "l", data = dweekend, 
     xlab = "time of the day (in hours)", ylab = "steps (average number)")
mtext("Weekend", side = 3, line = -1)
mtext("Average number of steps taken during the day (5 min interval)", 
      side = 3, outer = T, line = -1, font = 2)
```

